Hydra Battles and Cie

### 2021-05-25 doc/part-hydras.tex
minor changes
corrections to the pdf
correction of typos

\begin{remark}
Some notations (mainly names of fast-growing functions) of our development may differ slightly from the litterature. Although this fact does not affect our proofs, we are preparing a future 
version where the names $F\_alpha$, $f\_alpha$, $H\_alpha$, etc., are fully consistent with the cited articles. 
\end{remark}
%\section{Introduction}

@@

\begin{definition}
Let $0<\alpha<\epsilon_0$ be any ordinal, and $s$ a finite sequence $\langle s_1, s_2, \dots, s_N\rangle$ of strictly positive natural numbers.

We say that $s$ is \emph{$\alpha$-large} if the sequence $\langle \alpha_0=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots \rangle$ leads to $0$.

We say that $s$ is \emph{minimally $\alpha$-large} (in short:
\emph{$\alpha$-mlarge}) if $s$ is $\alpha$-large 
 and every strict prefix of $s$ leads to a non-zero ordinal (\emph{cf} Sect.~\vref{sect:path-to-def}).

@@

The following function ``gnaws'' an ordinal $\alpha$, following a sequence of indices (ignoring the $0$s).
\noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Paths.html\#gnaw}{Epsilon0.Paths}}

@@

Let us consider two integers $k$ and $l$, such that $0<k<l$. In order to check whether the interval $[k,l]$ is minimally large for $\alpha$, it is enough to
follow from $\alpha$ the path associated with the interval $[k,l($ and verify that the last ordinal we obtain is equal to $1$.

@@

Now, consider any natural number $k>0$. We would like to compute
a number $l$ such that the interval $[k,l]$ is $\alpha$-mlarge. So, 
the standard battle starting with $\iota(\alpha)$ and the replication factor $k$ will end after $(l-k+1)$ steps.

@@

\begin{Coqbad}
Function L (alpha: E0) (i:nat) :  nat := ...
\end{Coqbad}

@@

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider any  function which associates maps  any strictly positive integer $k$ to the number $l$, where 
the interval $[k,l)$ is $\alpha$-mlarge.

@@

@@

Let us now prove properties of any function $f$ (if any) which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plug-in. Moreover, they will give us some examples of
$L_\alpha$ for small values of $\alpha$.

Our exploration of the $L_\alpha$s  follows the usual scheme : transfinite induction, and proof-by-cases : zero, successors and limit ordinals.

@@

Let $\lambda<\epsilon_0$ be any limit ordinal. In a similar way as for successors, we decompose any path from $\lambda$  into a first step to
$\canonseq{\lambda}{k}$, followed by a path to $0$. In the following section, we assume that there exists à correct function for $\canonseq{\lambda}{k}$, \emph{for any strictly positive $k$}.

@@

We would like to get exact formulas for the ordinal $\omega^2$, a.k.a.
$\phi_0(2)$. This ordinal is the limit of the sequence $\omega\times i\;(i \in \mathbb{N}$. Thus, we have to study ordinals of this form, then use 
our lemma on limits.

@@

To this end, we prove a generic lemma, which expresses $L_{\omega^\alpha\times i}$ as an iterate of $L_{\omega^\alpha}$. Note that in this lemma, we assume that the fonction associated with $\alpha$ is stritly monotonous and 
greater or equal than the successor function, and prove that $L_{\omega^\alpha\times i}$satisfies  the same properties.

@@

\subsubsection{Back to hydra battles}

@@

\begin{Coqsrc}
Lemma H'_Omega_term (alpha : E0)  :
alpha <> Zero -> forall i k, 
  H'_ (Omega_term alpha i) k = iterate  (H'_ (Phi0 alpha)) (S i) k.
\end{Coqsrc}

@@

>>>>

\begin{remark}
Some notations (mainly names of fast-growing functions) of our development may differ slightly from the litterature. Although this fact does not affect our proofs, we are preparing a future version where the names $F\_alpha$, $f\_alpha$, $H\_alpha$, etc., are fully consistent with the cited articles.

\end{remark}
%\section{Introduction}

@@

\begin{definition}
Let $0<\alpha<\epsilon_0$ be any ordinal, and $s=\langle s_1, s_2, \dots, s_N\rangle$ a finite sequence of strictly positive natural numbers.

We say that $s$ is \emph{$\alpha$-large} if the sequence $\langle \alpha_0=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots \rangle$ leads to $0$. 
We say also that $s$ is \emph{minimally $\alpha$-large} (in short:
\emph{$\alpha$-mlarge}) if $s$ is $\alpha$-large 
 and every strict prefix of $s$ leads to a non-zero ordinal (\emph{cf} Sect.~\vref{sect:path-to-def}).

@@

The following function ``gnaws'' an ordinal $\alpha$, following a sequence of indices (ignoring the $0$s).

\noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Paths.html\#gnaw}{Epsilon0.Paths}}

@@

Let us consider two integers $k$ and $l$, such that $0<k<l$. In order to check whether the interval $[k,l]$ is minimally large for $\alpha$, it is enough to
follow from $\alpha$ the path associated with the interval $[k,l)$ and verify that the last ordinal we obtain is equal to $1$.

@@

Now, consider any natural number $k>0$ any ordinal $0<\alpha<\epsilon_0$.  We would like to compute
a number $l$ such that the interval $[k,l]$ is $\alpha$-mlarge. So, 
the standard battle starting with $\iota(\alpha)$ and the replication factor $k$ will end after $(l-k+1)$ steps.

@@

\begin{Coqbad}
Function L_ (alpha: E0) (i:nat) :  nat := ...
\end{Coqbad}

@@

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider any  function which  maps  any strictly positive integer $k$ to the number $l$, where 
the interval $[k,l)$ is $\alpha$-mlarge.

@@

\begin{todo}
 Check if the functions $L_\alpha$ are the same as
\cite{KS81}' functions $f\_alpha$ (p. 297).
\end{todo}

@@

Let us now prove properties of any function $f$ (if any) which satisfies 
\texttt{L\_spec}. We are looking for properties which could be used for writing \emph{equations} and prove the correctness of the function generated by the \texttt{coq-equations} plug-in. Moreover, they will give us some examples (for small values of $\alpha$).

Our exploration of the $L_\alpha$\,s  follows the usual scheme : transfinite induction, and proof-by-cases : zero, successors and limit ordinals.

@@

Let $\lambda<\epsilon_0$ be any limit ordinal. In a similar way as for successors, we decompose any path from $\lambda$  into a first step to
$\canonseq{\lambda}{k}$, followed by a path to $0$. In the following section, we assume that there exists a correct function computing  $L_{\canonseq{\lambda}{k}}$ for any strictly positive $k$.

@@

We would like to get exact formulas for the ordinal $\omega^2$, a.k.a.
$\phi_0(2)$. This ordinal is the limit of the sequence $\omega\times i\;(i \in \mathbb{N})$. Thus, we have to study ordinals of this form, then use 
our lemma on limits.

@@

To this end, we prove a generic lemma, which expresses $L_{\omega^\alpha\times i}$ as an iterate of $L_{\omega^\alpha}$. Note that in this lemma, we assume that the fonction associated with $\alpha$ is stritly monotonous and 
greater or equal than the successor function, and prove that $L_{\omega^\alpha\times i}$ satisfies  the same properties.

@@

\subsubsection{Back to hydra battles}
\label{def:L-alpha}

@@

\begin{Coqsrc}
Lemma H'_Omega_term (alpha : E0):
alpha <> Zero -> forall i k, 
  H'_ (Omega_term alpha i) k = iterate (H'_ (Phi0 alpha)) (S i) k.
\end{Coqsrc}

@@

\section{Conclusion}

In Sect.~\vref{sect:battle-length-notPR}, we prove that the length of hydra-battles (for a given hydra, according to the initial replication factor) is not primitive recursive in general. 
This proof uses properties of the Ackermann function, and the $H'_\alpha$, $F_\alpha$, $L_\alpha$ families of functions.
